---
title: "Environmental Statistics Final"
description: |
  A short description of the post.
author:
  - name: Halina Do-Linh
date: 11-14-2021
output:
  distill::distill_article:
    self_contained: false
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
```

```{r loading packages, include=FALSE}
library(tidyverse)
library(here)
library(lintr)
library(janitor)
library(lubridate)
```

# Technical Blog Guidelines

**Technical blog post [Due date: 12/2, 5pm]**

This blog post is a write up summarizing in text and with figures and/or tables your question, the data you have collected, your analysis plan, and your results. Your target audience should be other quantitative scientists and practitioners familiar with the basics of statistics and data science, but not necessarily experts in environmental science or the details of the methods studied in this course.

**Please email me your blog post in PDF format and as a live link by 5pm on December 2.**

Some guidelines for the blog post:

-   3-5 pages in length, including figures, tables, captions, and references list
-   1-3 tables or figures, each carefully labeled and captioned so that they are easily interpretable
-   Include scientific references when applicable
-   Include links to the underlying data you use. If your data cannot be shared publicly, note this in a short "data availability" statement at the end of your post.
-   [Recommended, but optional] If you can, include a link to a repository with your replication code. I will not evaluate your code as part of your grade, but this is good practice for reproducibility and transparency and will make your blog post more exciting to the outside world.

# General Guidelines

-   Motivate your question.

    -   Why is this important?

    -   Is there existing evidence on this question?

    -    If so, why is it inconclusive? If not, why not?
    
    


-   Describe your data.

    -    Where did you access it? What are its spatial and temporal features?

    -   What are its limitations?

    -   What do you know about the sampling strategy and what biases that may introduce?

    -   If helpful, you can use a histogram, scatterplot, or summary statistics table to describe your data.

-   Clearly describe your analysis plan.

    -   What is your analysis plan?

    -   Why did you choose this analysis, given your data and question?

    -   What are the limitations?

-   Summarize your results visually and in words.

    -   Show us your results in figure(s) and/or table(s) that are carefully labeled and captioned.

    -   Describe in the text (and orally when presenting) what you found, and how these results either do or do not help you answer your question.
    
    
```{r}
wells <- read_csv(here::here("_posts", "2021-11-14-stats-final", "data", "08_CA_AK.csv"))
```

```{r}
names(wells)
```

```{r}
wells_clean <- wells %>% 
  janitor::clean_names()

names(wells_clean)
```


```{r}
wells_county_depth <- wells_clean %>% 
  select("well_id", "county", "purpose_5", "date_completed", "depth", "screen_depth", "static_depth", "activity_type", "depth_m", "depth", "date", "depth_map")
```

```{r}
unique(wells_county_depth$county)
```

**Added region codes to counties:**

```{r}
region1 <- c("Butte", "Colusa", "El Dorado", "Glenn", "Lassen", "Modoc", "Nevada", "Placer", "Plumas", "Sacramento", "Shasta", "Sierra", "Siskiyou", "Sutter", "Tehama", "Yolo", "Yuba")
region1_code <- 1

region2 <- c("Del Norte", "Humboldt", "Lake", "Mendocino", "Napa", "Sonoma", "Trinity")
region2_code <- 2

region3 <- c("Almeda", "Contra Costa", "Marin", "San Francisco", "San Mateo", "Santa Clara", "Solano")
region3_code <- 3

region4 <- c("Alpine", "Amador", "Calaveras", "Madera", "Mariposa", "Merced", "Mono", "San Joaquin", "Stanislaus", "Tuolumne")
region4_code <- 4

region5 <- c("Monterey", "San Benitio", "San Luis Obispo", "Santa Barbara", "Santa Cruz", "Ventura")
region5_code <- 5

region6 <- c("Fresno", "Inyo", "Kern", "Kings", "Tulare")
region6_code <- 6

region7 <- c("Riverside", "San Bernardino")
region7_code <- 7

region8 <- "Los Angeles"
region8_code <- 8

region9 <- "Orange"
region9_code <- 9

region10<- c("Imperial", "San Diego")
region10_code <- 10

wells_region <- wells_county_depth %>% 
  mutate(region_code = case_when(county %in% region1 ~ region1_code,
                                 county %in% region2 ~ region2_code,
                                 county %in% region3 ~ region3_code,
                                 county %in% region4 ~ region4_code,
                                 county %in% region5 ~ region5_code,
                                 county %in% region6 ~ region6_code,
                                 county %in% region7 ~ region7_code,
                                 county %in% region8 ~ region8_code,
                                 county %in% region9 ~ region9_code,
                                 county %in% region10 ~ region10_code,
                                 TRUE ~ 0))
```

**Plot to see distribution of well depth (m). Had to take the log of well depth:**

```{r}
log_depth_m <- log(wells_region$depth_m)

depth_hist <- ggplot(data = wells_region, aes(log_depth_m)) +
  geom_histogram()

depth_hist
```

**Plot of depth(m) and region:**

```{r}
# to color cannot use "color" in histogram, need to use "fill"
# region_code originally has class numeric but needs to be class factor to fill

region_hist <- ggplot(data = wells_region, aes(x = region_code, fill = as.factor(region_code))) + 
  geom_histogram() 

region_hist
```


**Plot time**

```{r}
# found there are some dates with HMY format using table(wells-region$date_completed)

wells_time <- wells_region %>% 
  select(well_id, date_completed) %>% 
  mutate(date_completed = na.omit())
  mutate(date_completed = lubridate::guess_formats(date_completed, orders = "mdy"))

wells_time <- wells_region %>%
  mutate(date_completed = lubridate::mdY(date_completed))

time_hist <- ggplot(data = wells_region, aes(log_depth_m)) +
  geom_histogram()

time_hist

wells_time2 <- wells_region %>% 
  mutate(date_completed = lubridate::mdy(date_completed))

wells_time3 <- wells_region %>% 
  group_by(date_completed) %>% 
  count()


wells_valid_dates <- wells_region[!is.na(strptime(wells_region$date_completed, format = c("%m/%d/%y", "%m/%d/%Y", "%m/%d/%y HMS"))),]


wells_valid_dates <- wells_valid_dates %>% 
  mutate(date_completed = lubridate::parse_date_time(date_completed, orders = c("mdy", "mdy HMS", "mdY"))) %>% 
  filter(date_completed >= as.POSIXct("1900-01-01") & date_completed < as.POSIXct("2021-11-26"))



```

```{r}
ggplot(data = wells_valid_dates, aes(x = date_completed, y = depth_m, color = region_code)) + 
  geom_point()

```
```{r, deleted}
# wells_time_only_year <- wells_region %>% 
#   mutate(date_completed = na.omit(date_completed)
#   mutate(date_completed = format(date_completed, format = "%m/%d/%Y")) %>% 
#   group_by(date_completed) %>% 
#   count() %>% 
#   na.omit
#   mutate(date_completed = lubridate::mdy(date_completed))
# 
# wells_time_only_year <- wells_region %>% 
#   group_by(date_completed) %>% 
#   count()
# 
# wells_time_na <- wells_region %>% 
#   drop_na(date_completed) %>% 
#   mutate(date_completed = lubridate::parse_date_time(date_completed, c("mdy", "mdy HMS", "mdY")))
# 
# # wells_valid_dates <- wells_valid_dates %>% 
# #   group_by(date_completed) %>% 
# #   count()
```



-   What might you do next?

    -   One short analysis cannot fully answer an interesting scientific question.

    -   If you had time to collect more data or conduct more analysis, what would help you answer this question better?

Distill is a publication format for scientific and technical writing, native to the web.

Learn more about using Distill at <https://rstudio.github.io/distill>.
